---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
---


## Loading and preprocessing the data

For loading the data, we  assume that the original `activity.zip` file is present in the current working directory - no further checks are performed. The data is loaded into a dataframe and the `date` variable is transformed into a *date* object:

```{r}
# Load dplyr package for data munging;
# use "suppressMessages" to avoid messy output
suppressMessages(require(dplyr))

unzip("activity.zip")
stepdata <- read.csv("activity.csv", stringsAsFactors = FALSE)

stepdata <- stepdata  %>%
              mutate(date = as.Date(date))
```


## What is mean total number of steps taken per day?

We first summarize the total number of steps taken per day, and produce a histogram
to show the distribution:

```{r}
# Load ggplot2 package for all visualizations;
# use "suppressMessages" to avoid messy output
suppressMessages(require(ggplot2))

stepsPerDay <- stepdata %>%
                  group_by(date) %>%
                  summarize(totalSteps = sum(steps, na.rm = TRUE))

hist <- ggplot(stepsPerDay, aes(totalSteps)) + 
        geom_histogram(binwidth=1000, fill="lightgray", color="black") +
        labs(title = "Histogram of Steps per Day",
             x = "Number of steps per day",
             y = "Frequency") +
        theme(plot.title = element_text(size=16, face="bold", vjust=2),
              axis.title.x = element_text(vjust=-0.35),
              axis.title.y = element_text(vjust=0.35),
              panel.grid.minor.x = element_blank(),
              panel.grid.major.x = element_blank()) +
        scale_x_continuous(breaks=seq(0, 22000, by = 2000)) +
        scale_y_continuous(breaks=seq(0, 10, by = 2))
hist

```

We then calculate the **mean** and **median** values for the number of steps taken
per day:

```{r}
meanSteps <- stepsPerDay %>%
                summarize(mean(totalSteps))
medianSteps <- stepsPerDay %>%
                summarize(median(totalSteps))
```

The resulting values are:

- Mean number of steps per day: `r round(meanSteps, 2)`
- Median number of steps per day: `r medianSteps`

## What is the average daily activity pattern?

To extend the analysis above, we now summarize the original dataset by the
`interval` variable, dropping the date information on the way and calculating
the mean of every 5-minute-interval to reveal an averaged daily pattern of
steps, visualized in a time series plot:

```{r}
dailyPattern <- stepdata %>%
                  group_by(interval) %>%
                  summarize(meanSteps = mean(steps, na.rm = TRUE))

patt <- ggplot(dailyPattern, aes(interval, meanSteps)) +
          geom_line(color="firebrick") +
          labs(title = "Daily Activity Pattern",
               x = "5-minute interval",
               y = "Average number of steps") +
          theme(plot.title = element_text(size=16, face="bold", vjust=2),
                axis.title.x = element_text(vjust=-0.35),
                axis.title.y = element_text(vjust=0.35),
                panel.grid.minor.x = element_blank(),
                panel.grid.major.x = element_blank()) +
          scale_x_continuous(breaks=seq(0, 2400, by = 400))
patt

```

We filter the daily pattern for the **maximum** of the averaged number of steps
across all days:

```{r}
maxInterval <- dailyPattern %>%
                  top_n(1, meanSteps)
```

The time interval with the maximum number of average steps across all days is
`r maxInterval$interval`.

## Imputing missing values

Going back to the original dataset, we report the number of `NA`s in the `steps`
variable of this dataset:

```{r}
rowsNA <- stepdata %>%
            filter(is.na(steps)) %>%
            count

```

The number of `NA` rows in the dataset is `r rowsNA`.

To impute missing values into the dataset, we assume that the daily activity
patterns are sustainable and that the `NA` values are introduced into the original
dataset by the person simply not wearing the monitoring device. If this is true,
the assumption that the person might still be following their general activity
patterns for that particular time of the day / interval seems reasonable.

```{r}
# Combine the original dataset and the dailyPattern dataset to have a
# "fallback" value available in case the original dataset contains NA,
# then decide whether to use the original or the fallback value. Finally,
# round the steps variable and get back to an integer type to reflect that only
# a full number of steps can be performed by an individual.
stepdataImputed <- stepdata %>%
                      left_join(dailyPattern, by = "interval") %>%
                      transmute(steps = ifelse(is.na(steps), meanSteps, steps),
                                date = date,
                                interval = interval) %>%
                      mutate(steps = as.integer(round(steps)))

```

On this new dataset, we essentially repeat the calculations performed in section
**What is mean total number of steps taken per day?** and display a revised
histogram on the new dataset:

```{r}
stepsPerDayImp <- stepdataImputed %>%
                    group_by(date) %>%
                    summarize(totalSteps = sum(steps, na.rm = TRUE))

histImp <- ggplot(stepsPerDayImp, aes(totalSteps)) + 
          geom_histogram(binwidth=1000, fill="firebrick", color="black") +
          labs(title = expression(atop(bold("Histogram of Steps per Day"),
                                       atop(italic("NA values imputed")))),
               x = "Number of steps per day",
               y = "Frequency") +
          theme(plot.title = element_text(size=16, vjust=-1),
                axis.title.x = element_text(vjust=-0.35),
                axis.title.y = element_text(vjust=0.35),
                panel.grid.minor.x = element_blank(),
                panel.grid.major.x = element_blank()) +
          scale_x_continuous(breaks=seq(0, 22000, by = 2000)) +
          scale_y_continuous(breaks=seq(0, 50, by = 2))
histImp
```

We again calculate the **mean** and **median** values for the number of steps taken
per day, now on the imputed dataset:

```{r}
meanStepsImp <- stepsPerDayImp %>%
                     summarize(mean(totalSteps))
medianStepsImp <- stepsPerDayImp %>%
                     summarize(median(totalSteps))
```

The resulting values are:

- Mean number of steps per day: `r options(scipen=3);round(meanStepsImp, 2)`, vs. originally
`r round(meanSteps, 2)` steps
- Median number of steps per day: `r medianStepsImp`, vs. originally
`r medianSteps` steps

As to be expected, both the mean and the median of the imputed dataset are
higher than in the original dataset. The fact that the `na.rm = TRUE` option in
summarizing the original dataset removes those intervals when summing the daily
steps introduces a skew to the dataset. Adding the imputed values to the dataset

1. increase the overall number of steps to be summarized, hence also increasing
the mean and median
2. reduce some of the skew / imbalance of the dataset which is also reflected in
the mean and median values getting closer to each other

Of course, whether or not our imputing strategy is actually "improving" the
dataset with regards to the real number of steps taken by the individual cannot
be fully concluded from the data available. However, the assumption that the 
real number of steps taken in all `NA` interval is not 0 but a greater number 
oriented towards the general activity pattern still seems very valid.

## Are there differences in activity patterns between weekdays and weekends?

We create a new variable `weekday` on the imputed dataset, initially assign it 
the names of the respective weekdays using the `weekdays()` function in R, and
then distinguish between weekends and weekdays based on the labels of the day.

To exctract the daily patterns, we essentially repeat the analysis steps from 
the **What is the average daily activity pattern?** section, but additionally 
group by the new `weekday` variable. We then provide the activity plot 
subdivided by weekday/weekend activity.


```{r}
# Set the local time format to english to ensure that the weekday labels
# ("Saturday", "Sunday") are consistent across localized installs of R
Sys.setlocale("LC_TIME", "C")

stepdataWeekday <- stepdataImputed %>%
                      mutate(weekday = weekdays(date)) %>%
                      mutate(weekday = ifelse(weekday %in% c("Saturday",
                                                             "Sunday"),
                                              "weekend",
                                              "weekday")) %>%
                      mutate(weekday = as.factor(weekday))

weekdayPattern <- stepdataWeekday %>%
                      group_by(interval, weekday) %>%
                      summarize(meanSteps = mean(steps))

pattWD <- ggplot(weekdayPattern, aes(interval, meanSteps)) +
            geom_line(color="darkblue") +
            facet_wrap(~ weekday, ncol = 1) +
            labs(title = "Daily Activity Pattern",
                 x = "5-minute interval",
                 y = "Average number of steps") +
            theme(plot.title = element_text(size=16, face="bold", vjust=2),
                  axis.title.x = element_text(vjust=-0.35),
                  axis.title.y = element_text(vjust=0.35),
                  panel.grid.minor.x = element_blank(),
                  panel.grid.major.x = element_blank()) +
            scale_x_continuous(breaks=seq(0, 2400, by = 400))
pattWD

```


